TacticToe: Learning the Game of Tactics

### 2018-01-19 tactictoe_jv.tex
adding examples of generated tactictoe scripts
proof search: wip

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
Experiments on the \holfour standard 
library show that TacticToe is able to re-prove 59\% of the 7088 theorems in less 
than 10 seconds compared to 29\% proved by the HolyHammer system using the E ATP in its auto-schedule mode.

@@

\subsection{A proof tree and solved node}

@@

Starting from the bottom node, we can construct descendant of this node by 
choosing a goal $g_i^0$ on this node and choosing tactics $t1,t2,\ldots$ to 
apply to this goal. Here the result produces new nodes each containing a list 
of goals to be proven. If one of those list where empty, the goal $g_i^0$ would 
be proved.

A node is solved when every of its goal are solved.
A goal is solved when every of its children node are solved.

\subsection{Guiding the proof search}

@@

@@

Although the training process in each strategy on its own is fair, the 
selection of the 
best strategy in Section~\ref{sec:sel_param} should also be considered as a 
learning 
process. To ensure the global fairness, the final experiments in 
Section~\ref{sec:exp_full} 
runs the best strategy on the full dataset which is about 10 times larger. The
performance is minimally better on this validation set.

@@

@@

@@

Original \holfour proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
REWRITE_TAC [IMAGE_SURJ] THEN
  DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
  BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC
  THENL [
  irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
    REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
    REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
    FULL_SIMP_TAC std_ss [THE_DEF],
  ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
    RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
      [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC [] ]
\end{lstlisting}

@@

\tactictoe proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}

@@

Original \holfour proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}

@@

\tactictoe proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}

If a discovered proof is reasonable short, fast and understandable, they could 
replace its original in the \holfour standard library.

@@

Future work:
  Promising techniques that could not be evaluated. 
  + Orthogonalization of theorems~\cite{ckju-jsc15}. 
  + Try harder on term predictions. \cite{latest version of sepia}.

@@

\todo{is copied (modified a bit)}
We proposed a new proof assistant automation technique which combines 
tactic-based proof search with machine learning tactic prediction
Its implementation,
\tactictoe, achieves an overall performance of 59\% theorems on most of the 
\holfour standard library surpassing \eprover with auto-schedule. Its 
effectiveness is especially visible on 
theories which use inductive data structures, specialized decision procedures, 
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts 
usually reveal the high-level structure of the proof. %For these reasons, 
We therefore believe that predicting ITP tactics based on the current goal 
features is a very reasonable approach to automatically guiding proof search, 
and that accurate predictions can be obtained by learning from the knowledge 
available in today's large formal proof corpora. 
%And it seems to get us 
%closer to the way how humans learn to do theorem proving.

>>>>

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
Experiments on the \holfour standard 
library show that TacticToe is able to re-prove 59\% of 7088 theorems in less 
than 10 seconds. This is to be compared with the 29\% proved by the ATP Eprover 
in auto-schedule mode with 128 lemmas predicted and translated by HolyHammer.

@@

\subsection{A proof tree with two branching factors}

@@

Starting from the bottom node, we can construct descendant of this node by 
choosing a goal $g_i^0$ on this node and choosing tactics $t1,t2,\ldots$ to 
apply to this goal. Here the result produces new nodes each containing a list 
of goals to be proven.

Supervised learning introduction.
\subsection{Prior tactic policy}

\subsection{Prior node evaluation}

\subsection{Node selection}

@@

\paragraph{Solving a goal and a node}
Describe the case where a proof for a goal in a node is found.
If one of those list where empty, the goal $g_i^0$ would 
be proved.
A node is solved when every of its goal are solved.

\subsection{End of the proof search}
The proof search stops when one of these 3 conditions is realized:
\item It found a proof (i.e. the root node is solved). In this case, 
the search returns a minimized and 
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It has saturated. 
There is no more tactics to be applied to any open goal. (todo: define open 
goal somewhere).

\item It exceeded a given time limit.

The second case happens very rarely (less than 10 times when evaluating 7088 
theorems) and only at the beginning of the library when the learned dataset is 
small. So in our experiments, we just give the number of proof found relative 
to the number of proof attempts.

@@

\paragraph{Dataset}

All top-level theorems in the standard library are considered except:
\begin{itemize}
\item the 619 calls to store\_thm that contains a let binding, these are 
typically 
harder for \tactictoe as it may not have seen the particular local declaration 
before and therefore would have to find an alternative proofs. We also do not 
learn from these proofs because they were explictly local.
\item the 1175 calls to save\_thm. These are typically mostly trivial and easy 
theorems.
The remaining 7088 top-level theorems are evaluated in the full scale 
evaluation and the level of difficulty of these problem is assessed by a 
comparison with \eprover.
\end{itemize}

Among these types of theorems, every tenth theorems of the first third of the 
library is evaluated during tuning, which give us a total of 273 theorems to 
train different parameters on.

\paragraph{Strategy selection}
Although the training process in each strategy on its own is fair, the 
selection of the 
best strategy in Section~\ref{sec:sel_param} should also be considered as a 
learning 
process. To ensure the global fairness, the final experiments in 
Section~\ref{sec:exp_full} 
runs the best strategy on the full dataset which is about 10 times larger. The
performance is minimally better on this validation set.

@@

Constant progress: 
  From 45 percent on the same dataset in 5 seconds 
       59 percent in december in 10 seconds
  Probably 65-70 percent en of January in 60 seconds.

The progress cannot be attributed only to the time increase but to testing and 
debugging of parameters.

@@

%Name: 218 SURJ_INJ_INV
%Original proof time: 0.002033

@@

Original \holfour proof script (2.03 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
REWRITE_TAC [IMAGE_SURJ] THEN
  DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
  BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC
  THENL [
  irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
    REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
    REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
    FULL_SIMP_TAC std_ss [THE_DEF],
  ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
    RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
      [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC [] ]
\end{lstlisting}

@@

\tactictoe proof script (50.85 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}

@@

Original \holfour proof script (62.69 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}

%Name: 224 LUPDATE_APPEND2
%Original proof time:

@@

\tactictoe proof script (17.16 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}

If a discovered proof is reasonable short, fast, maintainable  and 
understandable, they could 
replace its original in the \holfour standard library. In the first case the 
tactictoe proof is slower but could still be considered becaues it is much more 
readable. In the second case, it is faster and more readable.

@@

Future work:
  Promising techniques that could not be evaluated. 
  + Orthogonalization of theorems~\cite{ckju-jsc15}. 
  + Try harder and smarted on term predictions. \cite{latest version of sepia}. 
  \cite{conjecturing}

More training than just orthogonalization of tactics.

@@

\todo{is copied (modified a bit)}
We proposed a new proof assistant automation technique which combines 
tactic-based proof search with machine learning tactic prediction
Its implementation,
\tactictoe, achieves an overall performance of 59\% theorems on 7088 theorems  
of the \holfour standard library surpassing \eprover with 
auto-schedule. Its 
effectiveness is especially visible on 
theories which use inductive data structures, specialized decision procedures, 
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts 
usually reveal the high-level structure of the proof. %For these reasons, 
We therefore believe that predicting ITP tactics based on the current goal 
features is a very reasonable approach to automatically guiding proof search, 
and that accurate predictions can be obtained by learning from the knowledge 
available in today's large formal proof corpora. 
%And it seems to get us 
%closer to the way how humans learn to do theorem proving.
